Functional Enrichment Tests: Intro and methods comparison

### 2023-06-21 DRAFT_Funct_Enrich.Rmd
knit changes
started addition of annot to FE

# Background

@@

Therefore, the quality of gene annotation is critical for accurate functional gene enrichment analysis. This involves careful curation of gene annotations based on experimental evidence from various sources, including transcriptomic and proteomic data, literature mining, and functional assays. This curation process can be particularly challenging for non-model species with limited genomic and transcriptomic resources. More information on this can be found at

- Example 1
- Example 2

@@

* Selection of appropriate gene background set. This set should not necessarily be the full set of genes in the genome. Rather it should be the set of genes expressed in the experiment. This is especially an issue with overrepresentation analysis and showed a consider impact on results when done incorrectly.
* Multiple hypothesis correction. A common metric used with enrichment results is a false discovery rate (FDR) correction. 
* Adequate methodological reporting including software and parameters used, versions of software and annotation files, all gene lists analyzed including subset and background sets, types of statistical tests and corrections performed, and any code used to conduct the analyses.

@@

* Differential expression
* Clustering
* WGCNA

@@

* The Gene Ontology (Thea)
* Annotation and filtration decisions (Sam) - include key terms
* Overrepresentation vs Enrichment (Thea)
* Explanation of statistics (Sam) - include key terms

@@

* Ontologizer (command line) (Thea)
* Panther (online GUI) - endorsed by The Gene Ontology (Thea)
* R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)

@@

## Fisher’s Exact Test

The Fisher’s exact test evaluates the relationship between two categorical variables: whether the ontology of a gene is represented in a test set or a background set. In the case of differential expression (DE), the test set is al DE genes and the background set is all genes included in DE analysis after filtering, not the entire transcriptome. The test asks whether the number of genes for a given GO term represented in a test set relative to the background is greater than what is expected due to chance alone (p).

@@

* $a =$ $n$ genes of GO term $i$ in both the test and background set
* $b =$ $n$ genes of GO $i$ in the background but not test
* $c =$ $n$ genes of GO $i$ absent from the test but present in the background
* $d =$ $n$ genes of GO $i$ absent in both the category and the background
* $n =$ $n$ genes in the background set

@@

The critical difference between Mann Whitney U and Fisher’s exact tests and are that Mann Whitney U’s permit the analysis of continuous rather than categorical variables. These continuous values can be log fold changes of differential expression or -log p-values associated with DE or inclusion in a WGCNA. Because it doesn’t rely on binary values denoting the existence of an effect, but rather the strength of an effect, the Mann Whitney U test is a more quantitative approach toward functional enrichment. The test asks whether the distribution of ranks for this continuous value in a GO term is significantly skewed from the distribution of ranks across all other GO terms. Like the background set employed in a Fisher’s exact test, “all other GO terms” refers to those terms included in genes used for analysis following filtering.

The skewedness of a GO term’s rank distribution relative to the background is represented by the statistic $U$. The p-value output by the Mann Whitney U test is calculated with the following equation:

@@

* $a =$ $n$ genes in GO term $i$
* $b =$ $n$ genes in the background set

@@

* The association file (-a) relating genes to GO terms (ends in .gaf for gene association file). A number of options are available through the annotations link on [this page](http://geneontology.org/docs/download-go-annotations/).
* The file (-g) containing the GO terminology and structure [go-basic.obo](http://geneontology.org/docs/download-ontology/).
* A population reference file (-p), which you will generate.
* Files representing your gene subsets of interest (-s). Multiple subsets can be provided at one time to Ontologizer and run in parallel by specifying a path to the directory containing only these files.

@@

Using the provided script located at pantherScore2.2/pantherScore2.2.pl, submit amino acid sequences to be searched against the PANTHER HMM library and assign appropriate IDs. Thus you may need to translate your transcript sequences using a tool such as TransDecoder. Searching your sequences against the PANTHER HMM library can be done on the command line. You will need to load perl and hmmer software packages. Test data are provided to check that everything installed properly. Once you are ready to go, you can run some version of the following. See command options for alternative variables to consider.

@@

    -l PANTER (l)ibrary with HMMs when using hmmsearch, binary PANTHER HMM model when using hmmscan
    -D display type for results
        options: I (interproscan), B (best hit), A (all hits)
    -o (o)utput file (redirect STDOUT)
    -i (i)nput fasta file to score
    -V (V)ery verbose (debug info to STDERR)

@@

Go to http://pantherdb.org. This will bring up the start page which will look something like this:

@@

* PANTHER Pathways
* PANTHER GO-Slim Molecular Function
* PANTHER GO-Slim Biological Process
* PANTHER GO-Slim Cellular Component
* PANTHER Protein Class

@@

* Visualizing results
* Interpreting results - gather consensus advice 
* R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)

@@

| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
| zzz | zzz | zzz | zzz | zzz | zzz | zzz |

@@

    ```{r setup, include = FALSE} `r ''`

@@

@@

# Run GOMWU and Fisher's exact tests on data of interest (see https://github.com/z0on/GO_MWU)

>>>>

@@

Therefore, the quality of gene annotation is critical for accurate functional gene enrichment analysis. This involves careful curation of gene annotations based on experimental evidence from various sources, including transcriptomic and proteomic data, literature mining, and functional assays. This curation process can be particularly challenging for non-model species with limited genomic and transcriptomic resources.

# Annotation

The following is a stepwise example or annotation of a gene set using Swiss-Prot such that Gene Ontology terms can be associated with each gene.

```{r engine='bash', eval=FALSE}
cd ../blastdb
curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
mv uniprot_sprot.fasta.gz uniprot_sprot_r2022_02.fasta.gz
gunzip -k uniprot_sprot_r2022_02.fasta.gz
```

```{r engine='bash', eval=FALSE}
/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
-in ../blastdb/uniprot_sprot_r2022_02.fasta \
-dbtype prot \
-out ../blastdb/uniprot_sprot_r2022_02
```

# Overview

@@

-   Selection of appropriate gene background set. This set should not necessarily be the full set of genes in the genome. Rather it should be the set of genes expressed in the experiment. This is especially an issue with overrepresentation analysis and showed a consider impact on results when done incorrectly.
-   Multiple hypothesis correction. A common metric used with enrichment results is a false discovery rate (FDR) correction.
-   Adequate methodological reporting including software and parameters used, versions of software and annotation files, all gene lists analyzed including subset and background sets, types of statistical tests and corrections performed, and any code used to conduct the analyses.

@@

-   Differential expression
-   Clustering
-   WGCNA

@@

-   The Gene Ontology (Thea)
-   Annotation and filtration decisions (Sam) - include key terms
-   Overrepresentation vs Enrichment (Thea)
-   Explanation of statistics (Sam) - include key terms

@@

-   Ontologizer (command line) (Thea)
-   Panther (online GUI) - endorsed by The Gene Ontology (Thea)
-   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)

@@

## Fisher's Exact Test

The Fisher's exact test evaluates the relationship between two categorical variables: whether the ontology of a gene is represented in a test set or a background set. In the case of differential expression (DE), the test set is al DE genes and the background set is all genes included in DE analysis after filtering, not the entire transcriptome. The test asks whether the number of genes for a given GO term represented in a test set relative to the background is greater than what is expected due to chance alone (p).

@@

-   $a =$ $n$ genes of GO term $i$ in both the test and background set
-   $b =$ $n$ genes of GO $i$ in the background but not test
-   $c =$ $n$ genes of GO $i$ absent from the test but present in the background
-   $d =$ $n$ genes of GO $i$ absent in both the category and the background
-   $n =$ $n$ genes in the background set

@@

The critical difference between Mann Whitney U and Fisher's exact tests and are that Mann Whitney U's permit the analysis of continuous rather than categorical variables. These continuous values can be log fold changes of differential expression or -log p-values associated with DE or inclusion in a WGCNA. Because it doesn't rely on binary values denoting the existence of an effect, but rather the strength of an effect, the Mann Whitney U test is a more quantitative approach toward functional enrichment. The test asks whether the distribution of ranks for this continuous value in a GO term is significantly skewed from the distribution of ranks across all other GO terms. Like the background set employed in a Fisher's exact test, "all other GO terms" refers to those terms included in genes used for analysis following filtering.

The skewedness of a GO term's rank distribution relative to the background is represented by the statistic $U$. The p-value output by the Mann Whitney U test is calculated with the following equation:

@@

-   $a =$ $n$ genes in GO term $i$
-   $b =$ $n$ genes in the background set

@@

-   The association file (-a) relating genes to GO terms (ends in .gaf for gene association file). A number of options are available through the annotations link on [this page](http://geneontology.org/docs/download-go-annotations/).
-   The file (-g) containing the GO terminology and structure [go-basic.obo](http://geneontology.org/docs/download-ontology/).
-   A population reference file (-p), which you will generate.
-   Files representing your gene subsets of interest (-s). Multiple subsets can be provided at one time to Ontologizer and run in parallel by specifying a path to the directory containing only these files.

@@

Using the provided script located at pantherScore2.2/pantherScore2.2.pl, submit amino acid sequences to be searched against the PANTHER HMM library and assign appropriate IDs. Thus you may need to translate your transcript sequences using a tool such as [TransDecoder](https://github.com/TransDecoder/TransDecoder/wiki). Searching your sequences against the PANTHER HMM library can be done on the command line. You will need to load perl and hmmer software packages. Test data are provided to check that everything installed properly. Once you are ready to go, you can run some version of the following. See command options for alternative variables to consider.

@@

```         
-l PANTER (l)ibrary with HMMs when using hmmsearch, binary PANTHER HMM model when using hmmscan
-D display type for results
    options: I (interproscan), B (best hit), A (all hits)
-o (o)utput file (redirect STDOUT)
-i (i)nput fasta file to score
-V (V)ery verbose (debug info to STDERR)
```

@@

Go to <http://pantherdb.org>. This will bring up the start page which will look something like this:

@@

-   PANTHER Pathways
-   PANTHER GO-Slim Molecular Function
-   PANTHER GO-Slim Biological Process
-   PANTHER GO-Slim Cellular Component
-   PANTHER Protein Class

@@

-   Visualizing results
-   Interpreting results - gather consensus advice
-   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)

@@

| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
|:-------|:-------|:-------|:-------|:-------|:-------|:-------|
| xxx    | xxx    | xxx    | xxx    | xxx    | xxx    | xxx    |
| yyy    | yyy    | yyy    | yyy    | yyy    | yyy    | yyy    |
| zzz    | zzz    | zzz    | zzz    | zzz    | zzz    | zzz    |

@@

````         
```{r setup, include = FALSE} `r ''`

@@

```
````

@@

# Run GOMWU and Fisher's exact tests on data of interest (see <https://github.com/z0on/GO_MWU>)
